---
title: "Lab 10: Filtering Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.height = 5)
knitr::opts_chunk$set(fig.width = 8.5)
knitr::opts_chunk$set(out.width = "100%")
knitr::opts_chunk$set(dpi = 300)

library(readr)
library(ggplot2)
library(dplyr)
library(viridis)
theme_set(theme_minimal())
```

## Pick a dataset

Today you will again have a choice of three datasets to work with. You can
choose either to investigate the daily team ELO ratings in the NBA:

```{r}
nba <- read_csv("https://statsmaths.github.io/stat_data/nba_elo_daily.csv")
```

Pick a dataset (feel free to delete the others from the Rmd file if you would
like), and work on the instructions in the next section.

## Instructions

This lab is similar to the last. You'll pick a dataset and then generate an
analysis of that data using data visualizations. The difference here is that
instead of a single plot, I would like you to construct 3-4 plots that
together tell a linear story. Each plot should be separated by a sentence or
two describing what the viewer should take away from the plot. Try to keep the
plots similar in some way; perhaps each simply highlights a different subset
of the data but has the same underlying layers.

Notice that each of the datasets for today are larger than you will probably
be able to use. You will need to filter the data to a particular team, year,
subset of curse words. You may further filter the data to highlight an even
smaller subset of the data within each plot.

## Analysis

The following plots show the relationship between the elo of the two teams that meet in the finals of the NBA in a selected year. This analysis specifically focuses on the rivalry between the Boston Celtics and the LA Lakers, and examines 4 years in which these two teams have met eachother in the finals. In every plot, the purple data points represent the LA Lakers and the green data points represent the Boston Celtics

The first plot shows the relationship between the elo of the Boston Celtics and the Minneapolis Lakers (later renamed the LA Lakers) during the 1959 season of the NBA. The Celtics prevailed over the Lakers in the Finals on April 9, 1959, beginning thier 8-year streak of NBA Finals wins. The Celtics swept the Lakers in the Finals (4-0), the first-ever sweep in NBA Finals history. This can be seen on the graph as the last four data points, where the two teams' elo values move in correlation with eachother. Overall, the Boston Celtics maintained a much higher elo than the Minneapolis Lakers for the entire 1959 NBA season.

```{r}

bos_mnl_1959 <- filter(nba, season == 1959, team %in% c("BOS", "MNL"))
bos_1959 <- filter(bos_mnl_1959, team == "BOS")
mnl_1959 <- filter(bos_mnl_1959, team == "MNL")

ggplot(bos_mnl_1959 , aes(date, elo)) + 
  ggtitle("NBA 1959 Season: Boston Celtics* vs. Minneapolis Lakers") +
  geom_line(color = "green", data = bos_1959) +
  geom_point(color = "green", data = bos_1959) + 
  geom_line(color = "purple", data = mnl_1959) +
  geom_point(color = "purple", data = mnl_1959)

```

The second plot shows the relationship between the elo of the Boston Celtics and the LA Lakers during the 1985 season of the NBA. The LA Lakers prevailed over the Boston Celtics in the Finals for the first time in NBA history. The two teams had very similar records throughout this season (LA Lakers 62-20 and the Boston Celtics 63-19). Unlike the previous graph, the Boston Celtics only began the season with a higher elo than the LA Lakers, later giving a lead to the LA Lakers about half-way into the season through their win in the Finals.

```{r}

bos_lal_1985 <- filter(nba, season == 1985, team %in% c("BOS", "LAL"))
bos_1985 <- filter(bos_lal_1985, team == "BOS")
lal_1985 <- filter(bos_lal_1985, team == "LAL")

ggplot(bos_lal_1985 , aes(date, elo)) + 
  ggtitle("NBA 1985 Season: Boston Celtics vs. Los Angeles Lakers*") +
  geom_line(color = "green", data = bos_1985) +
  geom_point(color = "green", data = bos_1985) + 
  geom_line(color = "purple", data = lal_1985) +
  geom_point(color = "purple", data = lal_1985)

```

The third plot shows the relationship between the elo of the Boston Celtics and the LA Lakers during the 2008 season of the NBA. The Boston Celtics won in the Finals over the LA Lakers on June 17, 2008, giving them a record 17 total national titles. This would mark the last national title to be won by the Boston Celtics to date, with one of the stars of the season, Paul Pierce, having his number retired by the Celtics organization on February 11, 2018. The elo of these two teams remained similar throughout the 2008 season, making the Finals a nail-biting experience and a devastating let-down for the LA Lakers.

```{r}

bos_lal_2008 <- filter(nba, season == 2008, team %in% c("BOS", "LAL"))
bos_2008 <- filter(bos_lal_2008, team == "BOS")
lal_2008 <- filter(bos_lal_2008, team == "LAL")

ggplot(bos_lal_2008 , aes(date, elo)) + 
  ggtitle("NBA 2008 Season: Boston Celtics* vs. LA Lakers") +
  geom_line(color = "green", data = bos_2008) + 
  geom_point(color = "green", data = bos_2008) + 
  geom_line(color = "purple", data = lal_2008) +
  geom_point(color = "purple", data = lal_2008)

```


The fourth graph shows the relationship between the elo of the Boston Celtics and the LA Lakers during the 2010 season of the NBA. The LA Lakers maintained a consistently larger elo than the Boston Celtics over the course of the season. The LA Lakers ultimtely prevailed over the Boston Celtics on June 17th in what would be the last meeting of the two teams in the NBA Finals until the present day.

```{r}
bos_lal_2010 <- filter(nba, season == 2010, team %in% c("BOS", "LAL"))
bos_2010 <- filter(bos_lal_2010, team == "BOS")
lal_2010 <- filter(bos_lal_2010, team == "LAL")

ggplot(bos_lal_2010 , aes(date, elo)) + 
  ggtitle("NBA 2010 Season: Boston Celtics vs. LA Lakers*") +
  geom_line(color = "green", data = bos_2010) + 
  geom_point(color = "green", data = bos_2010) + 
  geom_line(color = "purple", data = lal_2010) +
  geom_point(color = "purple", data = lal_2010)

```
