---
title: "Lab 21: Sixteen Personality Factors I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.height = 5)
knitr::opts_chunk$set(fig.width = 8.5)
knitr::opts_chunk$set(out.width = "100%")
knitr::opts_chunk$set(dpi = 300)

library(readr)
library(ggplot2)
library(dplyr)
library(viridis)
library(forcats)
library(smodels)
theme_set(theme_minimal())
```

## Instructions

Below you will find several empty R code scripts and answer prompts. Your task
is to fill in the required code snippets and answer the corresponding
questions.

## Sixteen Personality Factors: Pick Your Trait

Over the next two classes we are going to be looking at survey response
data from the Sixteen Personality Factor Questionnaire in order to practice
our skills at statistical inference:

```{r}
pf <- read_csv("https://statsmaths.github.io/stat_data/cattell_16pf.csv")
```

The dataset use the following fields:

- age: respondent's age in years
- gender: respondent's self-selected gender
- country: two letter IATA code for the respondent's IP
- elapsed: time taken to complete quiz in seconds
- warmth: personality score from 1-20
- reasoning: personality score from 1-20
- emotional_stability: personality score from 1-20
- dominance: personality score from 1-20
- liveliness: personality score from 1-20
- rule_consciousness: personality score from 1-20
- social_boldness: personality score from 1-20
- sensitivity: personality score from 1-20
- vigilance: personality score from 1-20
- abstractedness: personality score from 1-20
- privateness: personality score from 1-20
- apprehension: personality score from 1-20
- openness_to_change: personality score from 1-20
- self_reliance: personality score from 1-20
- perfectionism: personality score from 1-20
- tension: personality score from 1-20
- baseline: average score across all 16 personality traits

*To start with, select a particular trait that you will use for the first
bank of questions. I suggest picking something that popped out when you took
the test. You can pick any of the 16 other than sensitivity.*

Produce a bar plot of the personality scores for your trait.

Describe the distribution. Why does a bar plot work here even though the
variable is numeric?

**Answer**:

Produce a confidence interval for the mean of your trait.

```{r}
model <- lm_basic(openness_to_change ~ 1, data = pf)
reg_table(model, level = 0.95)
```

Now, produce a dataset that consists only of responses from the country of
Hungary. The 2-letter country code for Hungary is "HU". Produce a confidence
interval for your trait on the Hungarian dataset.

```{r}
hu <- filter(pf, country == "HU")
model <- lm_basic(openness_to_change ~ 1, data = hu)
reg_table(model, level = 0.95)

```

What do you notice about this confidence interval compared to the original
one? Can you explain why this is the case?

**Answer**: The confidence interval for the data of those in Hungary only has a much larger confidence interval than that of the whole dataset. This might be explained by the fact that there is less data points in the set of responses from Hungary than there are in the complete dataset, making it more difficult to find a distinct/small interval of confidence.

Find a 95% confidence interval for the difference between the average
male and female value for your personality trait on the Hungarian data.


**Answer**: 

```{r}
model <- lm_basic(openness_to_change ~ 1 + gender, data = hu)
reg_table(model, level = 0.95)

```


Is there a statistically significant difference between men and women's
average score? If so, what direction is this difference? Does this challenge
or confirm traditional gender stereotypes (note: not all personality traits
have one)?

**Answer**: No, there is not a statistically significant difference between men and women's average score for openness to change, because the confidence interval includes the value 0. I am not aware of a gender stereotype that states whether men or women have a higher degree of openness to change, so I cannot say if this challenges or confirms any existing stereotype.

Construct a new dataset that only has ages from 30-49 (from the original `pf`,
not just the Hungarian subset). Hint: you can use the filter function twice.

```{r}
age30thru49 <- filter(pf, age >= 30, age <= 49)
model <- lm_basic(openness_to_change ~ 1, data = age30thru49)
reg_table(model, level = 0.95)
```

On the dataset of people ages 30 to 49, create a variable called `fourties` if
age is greater than or equal to 40.


Fit a regression on with your personality trait as a response to test for the
difference between its mean for people in their 30s versus people in their
40s. Compute a confidence interval for this difference.

```{r}
age30thru49$fourties <- c(age30thru49$age >= 40)

```

Is there a statistically significant difference between 30s and 40s average
scores? If so, what direction is this difference? Does this challenge or
confirm traditional age stereotypes (note: not all personality traits
have one)?

```{r}
model <- lm_basic(openness_to_change ~ 1 + fourties, data = age30thru49)
reg_table(model, level = 0.95)
```
**Answer**: Yes, there is a statistically significant difference between 30s and 40s average openness to change because the confidence interval does not include 0 - however, the lower bound is very near 0. I would assume that the older the person is, the less open to change they would be - especially in terms of politics and religion. This result challenges this assumption, because it says that those who are in their fourties are slightly more open to change than those in their 30's.

Draw a bar plot of the variable `country` over the whole dataset. Take
note of the number of countries with a very small number of responses.

```{r}
ggplot(pf, aes(country)) + 
  geom_bar()

```

Fitting a model with all of these countries is possible but not very
useful. When we only have a few responses from some places it makes those
few regions with a lot of data harding to identify and analyze. Fortunately
there is a solution in the function `fct_lump`.

Fit a regression model on the entire dataset by the country variable lumped
into 5 categories.

```{r}
model <- lm_basic(openness_to_change ~ 1 + fct_lump(country, 5), data = pf)
reg_table(model, level = 0.95)
```

Using the previous model, which country (not including "Other") has the
highest score for your trait? Which has the lowest?

**Answer**: IN (India) has the highest score for openness to change, while GB (Great Britain) has the lowest average score for openness to change.

## Fitting Sensitivity Scores

Fit a model that predicts sensitivity as a function of the baseline
score.

```{r}
model <- lm_basic(sensitivity ~ 1 + baseline, data = pf)
reg_table(model, level = 0.95)
```

Is the slope statistically significantly different from 1? Why is this an
interesting question in the context of the data?

**Answer**: Yes, because the confidence interval does not include 1. This is interesting, because it means that a persons sensitivity is generally lower than what is expected from the persons baseline score.

Fit a regression model that uses both gender and the baseline score to
predict your trait.

```{r}
model <- lm_basic(sensitivity ~ 1 + baseline + gender, data = pf)
reg_table(model, level = 0.95)
```

Describe the slope for the gender term in words:

**Answer**:

Add predictions from the previous model to `pf`.

```{r}
model_pred <- add_prediction(pf, model)
```

Plot the baseline score as a function of `model_pred`, coloring the points
base on the gender variable.

```{r}
ggplot(model_pred, aes(gender, baseline)) + 
  geom_point(aes(color = gender))
```

What do the predicted values look like?

**Answer**: They appear to be generally the same between male and female.


